Implementing Distributed Systems Directly\\with I/O Automata in ioa++

### 2011-09-30 papers/paper1/design.tex
New case study and related work.

ioa++ is an implementation of the component model described in Section~\ref{component_model} for POSIX environments in C++.
Our implementation provides services for executing local actions and facilities for manipulating the set of automata and bindings.
The architecture of our framework is depicted in Figure~\ref{framework_architecture}.
The Scheduler is responsible for selecting and executing local actions subject to the I/O automata model and the concurrency constraints described in Section~\ref{component_model}.
All user automata inherit from a common base class that provides facilities for asynchronously creating and destroying child automata and bindings.
The services provided by the base class are indicated by the Base Automaton layer shown in Figure~\ref{framework_architecture}.
The Base Automaton class contains statically composed System Actions that are used to communicate with the System Automaton.
In most situations, the services provided by the Base Automaton class are too low-level to be used directly by a User Automaton.
An Automaton Manager presents a high-level interface for asynchronously managing a single child automaton using the services of the Base Automaton class.
Similarly, a Binding Manager presents a high-level interface for asynchronously managing a single binding.

@@

The controller encodes a \emph{policy} that selects the next local action.
For the framework to be an valid implementation of the I/O automata model, the controller used by the scheduler should be \emph{fair} as defined in Section~\ref{component_model}.
The run-time system can be initialized with different controllers to achieve different performance objectives.

\paragraph*{Explicit dynamic scheduling}
The I/O automata model assumes that the scheduler knows about every local action in the system.
There are two approaches for achieving this in practice.
In the first approach, automata declare \emph{statically} all of their actions to the scheduler.
The controller then has the responsibility of cycling through the actions according to some fair policy.
Since the scheduler has no knowledge about what actions are enabled, i.e., have a precondition that is true, this approach reduces to trying each action with brute force.

The approach used in our framework is to allow automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
The controller then has the responsibility of processing the set of requested actions according to some fair policy.
Let an action be \emph{enabled} if its precondition is true and \emph{disabled} otherwise.
Observe that executing an action potentially changes the state of the automaton and causes the set of enabled actions to change.
Consequently, an automaton is given a opportunity to declare actions to the scheduler after one of its actions is executed.
Since the state and therefore preconditions of all automata are independent, each automaton is responsible for scheduling its own actions.
For bootstrapping, each automaton is allowed to schedule actions when it is created.

In this approach, scheduling disabled actions is unnecessary.
Let $a$ be a scheduled disabled action.
In order for $a$ to become enabled, the automaton associated with the action must change state by executing an enabled action $b$ sometime after $a$ is scheduled.
Consequently, the scheduling of the disabled action $a$ can be deferred until the execution of action $b$.
Notice that refraining from scheduling disabled actions does not affect the fairness of the scheduler since they cannot cause a state change when selected.
This result leads to an programming idiom where the precondition of each action is checked before it is added to the schedule.

In contrast, an automaton should ensure that the scheduler will consider all actions that are currently enabled.
A simple technique for accomplishing this is to schedule all enabled actions.
More efficient schemes are possible if one considers how an action might enable other actions and which actions have already been scheduled.
Not declaring an enabled action violates the fairness of the controller since it will never select an action that may cause a state change.
Not scheduling an (enabled) action is a common mistake.

\paragraph*{Binding predicates and scheduling}
The preceding discussion on scheduling assumes that an action in one automaton cannot affect preconditions in another automaton.
This is not entirely correct when one considers the system automaton and binding predicates.
For example, imagine an automaton with a single output action whose only precondition is that it must be bound and assume that the automaton only schedules the action when enabled.
When the automaton is created, its only action is disabled.
Consequently, it will not schedule its action and therefore will not be executed.
Suppose that some time after the automaton is created, another automaton successfully binds to the output action.
The bind changes the precondition but the automaton will never be executed since it has no way of scheduling the output action.
While this example may seem contrived, it is representative of situations where an automaton is waiting for one or more of its outputs to be bound: a situation that occurs quite often in practice.
To remedy this, our framework takes advantage of the fair scheduling criteria of the I/O automata model and schedules an output action after a successful bind or unbind.

\paragraph*{Actions of no-longer-existing automata}
Every action in the set of actions considered by the controller is scheduled by its associated automaton.
Since automata can be destroyed, an action in the set of actions owned by the controller might belong to an automaton that no longer exists.
Our solution is to check if the automaton exists before executing an action.
This check is performed by the dispatcher and avoids the overhead associated with synchronizing the set of actions with the set of automata every time an automaton is destroyed.
Our solution assigns each automaton a unique identifier called an \emph{automaton id (aid)} and that the set of actions in the scheduler turns over comparatively faster than the set of aids.

\paragraph*{Delayed actions}
Scheduling activities for a future time is a useful feature for a real system because it allows the system to sleep between activities.
The framework should permit the development of automata that act as timers and alarms.
The corresponding feature that must be supported by the framework is scheduling an action at a future time.
Conceptually, the action is held until the time associated with the action at which point it is added to the set of actions being considered by the scheduler.
Actions scheduled in the past or present are released immediately to the scheduler.
When the same action is scheduled at two different times, the earliest time is used.
% Things I didn't say:  No canceling.

\paragraph*{File descriptor events}
File descriptors are used to communicate with the outside world in a POSIX environment.
Observe that there is no concept of blocking in the I/O automata model and introducing blocking I/O operations into action effects adversely affects the asynchronous and concurrent nature of I/O automata.
Thus, we require techniques that allow a program (1) to wait until a file descriptor is ready for I/O and (2) to perform non-blocking I/O operations.
This implies the use of either the Proactor or Reactor pattern~\cite{schmidt2000pattern}.
For our framework, we used the Reactor pattern since (1)~the Reactor pattern is simpler and (2) all major operating systems have non-blocking I/O and some kind of synchronous I/O multiplexing whereas asynchronous I/O is much less standardized.

Our implementation of the Reactor pattern causes an action to be scheduled when a file descriptor is ready for reading or writing.
A user automaton creates a file descriptor and configures it for non-blocking I/O.
The automaton then tells the scheduler to release an action (typically an internal action) whenever the automaton becomes ready for reading (or writing).
Subsequent requests made using the same file descriptor and event (ready for reading, ready for writing) are ignored.
Once the file descriptor is ready, the action is released for consideration by the controller.
The scheduler also provides a special function for closing file descriptors that purges them from the reactor.

\paragraph*{Global FIFO scheduler}
The scheduler included with the framework is single-threaded and uses a single queue of actions that is processed using first-in/first-out semantics.
Scheduling is idempotent to prevent unbounded growth of the queue.
The global FIFO scheduler uses the Reactor pattern to implement delayed actions and file descriptor events.

@@

\paragraph*{System action protocol}
Our approach to system actions follows the model introduced in Section~\ref{dynamics}.
All automata are statically composed with a system automaton via system actions for dynamically creating child automata, binding external actions, unbinding external actions, and destroying child automata.
The system automaton implements a request-response protocol where an automaton requests a system action and the system automaton processes the request and responds with either success or failure.
Associated with each request is a \emph{subject} which is the automaton's local name for an automaton or binding.
Subjects allow an automaton to have multiple outstanding requests.

Automata use a simple protocol for creating and destroying child automata.
Each child automaton is represented by a unique subject which is in one or two of five states: CreateSend, CreateRecv, CreateDone, DestroySend, and DestroyRecv.
Subjects start in the CreateSend state which indicates that the request for this subject is waiting to be sent to the system automaton.
Once the request has been sent, the subject transitions to the CreateRecv state and waits for a response from the system automaton.
Once the system automaton sends the response, the subject transitions to the CreateDone state.
When the user automaton requests that the child automaton be destroyed, the subject is either (1) added to the DestroySend state if the subject is in CreateRecv or CreateDone or (2) removed from all states if the subject is in CreateSend.
A subject appearing in both CreateDone and DestroySend transitions to the DestroyRecv state when the automaton makes the request to destroy.
The system automaton can process system actions in any order.
This protocol ensures that an automaton doesn't request that a child automaton be destroyed until it receives confirmation that it was created.
The subject is forgotten when the system automaton responds with the result of the destroy.

The protocol for binding is similar but addresses the possibility that a binding can be dissolved at any point due to the asynchronous destruction of automata.
Define analogous states for the binding state machine: BindSend, BindRecv, BindDone, UnbindSend, and UnbindRecv.
A binding subject can be in either BindRecv, BindDone, UnbindSend, or UnbindRecv when it receives a result from the system automaton indicating that the binding was dissolved.

\paragraph*{Automaton and Binding managers}
The protocol used by automata to asynchronously manage child automata and bindings is suitable for simple applications but becomes unwieldy when creating complicated constellations.
The Automaton Managers and Binding Managers provide an individualized facade to the system action state machines.
Creating an Automaton Manager initiates the creation of a new child automaton and invoking a destroy method initiates the destruction of a child automaton.
Automaton Managers can be observed using the Observer~\cite{gamma1995design} pattern to determine when and how the creation/destruction process succeeds.
Binding Managers are similar except they do not begin the process of binding until the automata providing the output action and input action have been created, thus, Binding Managers observe Automaton Managers.

>>>>

ioa++ implements the component model described in Section~\ref{component_model}, for POSIX environments (in C++).
Our implementation provides services for executing local actions and facilities for manipulating the set of automata and bindings.
The architecture of the ioa++ framework is depicted in Figure~\ref{framework_architecture}.
The Scheduler is responsible for selecting and executing local actions subject to the I/O automata model and the concurrency constraints described in Section~\ref{component_model}.
All user automata inherit from a common base class that provides facilities for asynchronously creating and destroying child automata and bindings.
The services provided by that common base class are indicated by the Base Automaton layer shown in Figure~\ref{framework_architecture}.
The Base Automaton class contains statically composed System Actions that are used to communicate with the System Automaton.
In most situations, the services provided by the Base Automaton class are too low-level to be used directly by a User Automaton.
An Automaton Manager presents a high-level interface for asynchronously managing a single child automaton using the services of the Base Automaton class.
Similarly, a Binding Manager presents a high-level interface for asynchronously managing a single binding.

@@

The controller encodes a \emph{policy} that selects the next local action.
For the framework to be an valid implementation of the I/O automata model, the controller used by the scheduler must be \emph{fair} as defined in Section~\ref{component_model}.
The run-time system can be initialized with different controllers to achieve different performance objectives.
Controllers can use multiple threads of execution to exploit the concurrency available in a system of automata.
The ioa++ framework is designed so that all native concurrency structures (e.g., threads, mutexes) are encapsulated in the dispatcher and controller.

\paragraph*{Explicit dynamic scheduling}
The I/O automata model assumes that the scheduler knows about every local action in the system.
There are two approaches for achieving this in practice.
In the first approach, automata declare \emph{statically} all of their actions to the scheduler.
The controller then has the responsibility of cycling through the actions according to some fair policy.
Since the scheduler has no knowledge about what actions are enabled, i.e., have a precondition that is true, this approach reduces to naively trying each action.

A more focused approached used in our framework is to allow automata to declare \emph{dynamically} the actions they would like the scheduler to consider.
The controller then has the responsibility of processing the set of requested actions according to some fair policy.
Let an action be \emph{enabled} if its precondition is true and \emph{disabled} otherwise.
Observe that executing an action potentially changes the state of the automaton and causes the set of enabled actions to change.
Consequently, an automaton is given the opportunity to declare actions to the scheduler after one of its actions is executed.
Since the state variables (and therefore preconditions) of all automata are independent, each automaton is responsible for scheduling its own actions.
For bootstrapping, each automaton is allowed to schedule actions when it is created.

In this approach, scheduling disabled actions is unnecessary.
Let $a$ be a scheduled disabled action.
In order for $a$ to become enabled, the automaton associated with the action must change state by executing an enabled action $b$ sometime after $a$ is scheduled.
Consequently, the scheduling of the disabled action $a$ can be deferred until the execution of action $b$.
Notice that refraining from scheduling disabled actions does not affect the fairness of the scheduler since they cannot cause a state change when selected.
This result leads to a guarded programming idiom where the precondition of each action is checked before it is added to the schedule.

However, an automaton should still ensure that the scheduler will \emph{consider} all actions that are currently enabled.
A simple technique for accomplishing this is to \emph{schedule} all enabled actions.
More efficient schemes are possible if one considers how an action might enable other actions and which actions have already been scheduled.
However, this must be done carefully.
Not declaring an enabled action would violate the fairness of the controller since it will never select an action that may cause a state change.
Not scheduling an (enabled) action could be another potential mistake.

\paragraph*{Binding predicates and scheduling}
%% This assumption is not implicit because we state that the preconditions are independent.
The preceding discussion on scheduling assumed that an action in one automaton cannot affect preconditions in another automaton.
This is not entirely correct when one considers the system automaton and binding predicates.
For example, imagine an automaton with a single output action whose only precondition is that it must be bound and assume that the automaton only schedules the action when enabled.
When the automaton is created, its only action is disabled.
Consequently, it will not schedule its action and therefore will not be executed.
Suppose that some time after the automaton is created, another automaton successfully binds to the output action.
The bind changes the precondition but the automaton will never be executed since it has no way of scheduling the output action.
While this example may seem contrived, it is representative of situations where an automaton is waiting for one or more of its outputs to be bound: a situation that occurs quite often in practice.
To remedy this, our framework takes advantage of the fair scheduling criteria of the I/O automata model and schedules an output action after each successful bind or unbind.

\paragraph*{Actions of no-longer-existing automata}
Every action in the set of actions considered by the controller is scheduled by its associated automaton.
Since automata can be destroyed, an action in the set owned by the controller might belong to an automaton that no longer exists.
Our solution is to check if the automaton exists before executing an action.
This check is performed by the dispatcher and avoids the overhead associated with synchronizing the set of actions with the set of automata every time an automaton is destroyed.
Our solution assigns each automaton a unique identifier called an \emph{automaton id (aid)}, assuming that automata turn over less frequently overall than the set of actions in the scheduler.

\paragraph*{Delayed actions}
Scheduling activities for a future time is a useful feature for a real system because it allows the system to sleep between activities, i.e., the framework should support the development of automata that act as timers and alarms.
Conceptually, an action is held until the time associated with it, at which point it is added to the set of actions being considered by the scheduler.
Actions scheduled in the past or present are released immediately to the scheduler.
When the same action is scheduled at two different times, the earliest time is used.
% Things I didn't say:  No canceling.

\paragraph*{File descriptor events}
File descriptors are used to communicate with the outside world in a POSIX environment.
Observe that there is no concept of blocking in the I/O automata model and introducing blocking I/O operations into action effects may adversely affect the asynchronous and concurrent nature of I/O automata.
Thus, we require techniques that allow a program (1) to wait until a file descriptor is asynchronously ready for I/O and (2) to perform non-blocking I/O operations.
This implies the use of either the Proactor or Reactor pattern~\cite{schmidt2000pattern}.
For our framework, we used the Reactor pattern since (1)~the Reactor pattern is simpler and (2) all major operating systems have non-blocking I/O and some kind of synchronous I/O multiplexing whereas asynchronous I/O is much less widespread.

Our use of the Reactor pattern causes an action to be scheduled when a file descriptor is ready for reading or writing.
A user automaton creates a file descriptor and configures it for non-blocking I/O.
The automaton then tells the scheduler to release an action (typically an internal action) whenever the automaton becomes ready for reading (or writing).
Subsequent requests made using the same file descriptor and event (ready for reading, ready for writing) are ignored.
Once the file descriptor is ready, the action is released for consideration by the controller.
The scheduler also provides a special function for closing file descriptors that purges them from the reactor.

\paragraph*{Global FIFO scheduler}
The basic scheduler included with the ioa++ framework distribution is single-threaded and uses a single queue of actions that is processed using first-in/first-out semantics.
Scheduling is idempotent to prevent unbounded growth of the queue.
The global FIFO scheduler uses the Reactor pattern to implement delayed actions and file descriptor events.

@@

Our approach to system actions follows the model introduced in Section~\ref{dynamics}.
All automata are statically composed with a system automaton via system actions for dynamically creating child automata, binding external actions, unbinding external actions, and destroying child automata.
The system automaton implements a request-response protocol where an automaton requests a system action and the system automaton processes the request and responds with an indication of either success or failure.
Associated with each request is a \emph{subject} which is the automaton's local name for an automaton or binding.
Subjects allow an automaton to have multiple outstanding requests.

Automata use a simple protocol for creating and destroying child automata.
Each child automaton is represented by a unique subject which is in one or two of five states: CreateSend, CreateRecv, CreateDone, DestroySend, and DestroyRecv.
Subjects start in the CreateSend state which indicates that the request for this subject is waiting to be sent to the system automaton.
Once the request has been sent, the subject transitions to the CreateRecv state and waits for a response from the system automaton.
Once the system automaton sends the response, the subject transitions to the CreateDone state.
When the user automaton requests that the child automaton be destroyed, the subject is either (1) added to the DestroySend state if the subject is in CreateRecv or CreateDone or (2) removed from all states if the subject is in CreateSend.
A subject appearing in both CreateDone and DestroySend transitions to the DestroyRecv state when the automaton makes the request to destroy it.
The system automaton can process system actions in any order.
This protocol ensures that an automaton doesn't request that a child automaton be destroyed until it receives confirmation that it was created.
The subject is forgotten when the system automaton responds with the result of the destroy action.

The protocol for binding is similar but addresses the possibility that a binding can be dissolved at any point due to the asynchronous destruction of automata.
We define analogous states for the binding state machine: BindSend, BindRecv, BindDone, UnbindSend, and UnbindRecv.
A binding subject can be in either BindRecv, BindDone, UnbindSend, or UnbindRecv when it receives a result from the system automaton indicating that the binding was dissolved.

\paragraph*{Automaton and Binding managers}
The protocol used by automata to asynchronously manage child automata and bindings is suitable for simple applications but becomes unwieldy when creating more complicated ones.
Automaton Managers and Binding Managers provide individualized facades to the system action state machines.
Creating an Automaton Manager initiates the creation of a new child automaton and invoking a destroy method initiates the destruction of a child automaton.
Automaton Managers can be monitored using the Observer~\cite{gamma1995design} pattern to determine when and how the creation/destruction process succeeds.
Binding Managers are similar except they do not begin the process of binding until the automata providing the output action and input action have been created: thus, Binding Managers observe Automaton Managers.

### 2011-09-30 papers/paper1/evaluation.tex
New case study and related work.

ioa++ permits independent actions to be executed concurrently.
The goal of this evaluation is (1) to show that concurrent execution is indeed possible and (2) to show that the degree of exploitable concurrency depends on the interactions of the I/O automata comprising the system and the overhead of the system.
To test concurrent execution, we execute a system whose actions can be configured to span the range from having no independent actions to having all independent actions.
We measure the time required to execute a fixed number of actions using a scheduler with a configurable number of threads.
We then calculate the speed-up by comparing an execution using one thread to an execution using two threads.
We also vary the complexity of each action to gain insight into how overhead and synchronization affect concurrent execution.

An automaton $R$ contains an input, output, and internal action.
The output and internal action execute a routine whose complexity is proportional to the parameter $N$.
The automaton executes a fixed number of local actions.
The automaton schedules the internal action with probability $\sqrt{\rho}$ and schedules the output action otherwise.
The system $S$ to be executed consists of two $R$ automata composed so the output action of one is composed with the input action of the other.
If we divide the execution into rounds where each automaton executes a single action, the probability that both $R$ automata execute an internal action is $\sqrt{\rho} \times \sqrt{\rho} = \rho$.
Thus, we can use the parameter $\rho$ to vary the independence of the automata.

For this experiment, we developed a simple multi-threaded scheduler that assigns each local action to a thread based on the automaton identifier of the local action.
When a single thread is used, all actions are executed by the same thread.
When two threads are used, one thread executes the actions of one $R$ automaton while the other thread executes the actions of the other $R$ automaton.

A trial consists of a choice for $\rho$, $N$, and the number of actions to be executed.
For the experiments, the parameter $\rho$ was varied from $0.0$ to $1.0$ in increments of $0.1$ and the parameter $N$ was varied from $1$ to $1000000$ by factors of $10$.
The number of actions executed by each automaton was fixed at 1000.
Each trial was repeated 1000 times.
All calculations were performed assuming a normal distribution at a confidence level of 95\%.

The driver for a single trial can be found in examples/random.cpp of the ioa++ package.
The trials were performed on a Mac G4 running OS X 10.5.8 with two 2.66 GHz Dual-Core Intel Xeon processors with 2GB of memory.
The code was compiled and linked with i686-apple-darwin9-g++-4.0.1 build 5493 with -O2 optimization and the -DPROFILE flag to enable profiling in ioa++.

@@

Figure~\ref{speed_up} shows the speed-up when system $S$ was executed with one and two threads.
Confidence intervals are shown with largest being 0.0165.
When $\rho = 0$, every action is a bound output action and therefore depends on both $R$ automata.
Consequently, every action must be serialized yielding a maximum speed-up of 1.
When $\rho = 1$, every action is an internal action and can be executed concurrently with a corresponding maximum speed-up of 2.

@@

Figures~\ref{overhead1} and \ref{overhead2} show the per thread average of the fraction of time devoted to framework code and synchronization calls for the single and multi-threaded executions.
Confidence intervals are shown with largest being 0.002155 for Figure~\ref{overhead1} and 0.001781 for Figure~\ref{overhead2}.
Since the number of actions to be executed is constant, the single threaded execution shows a relatively consistent fraction for each value of $N$.
The fraction decreases as $\rho$ goes to 1 and the system transitions from acquiring two locks to one lock.
The multi-threaded execution shows that the fraction of time devoted to system overhead decreases as $\rho$ goes to 1 and is more pronounced when $N$ is large.
For example, when $N=1000000$ one thread spends half of its time waiting on the other thread when $\rho = 0$ and spends virtually no time on synchronization when $\rho = 1$.

This experiment indicates that concurrent execution is possible with ioa++ but a speed-up will only be achieved if (1) enough independent actions are enabled and (2) the duration of the independent actions overcome the overhead of ioa++.
The overhead of ioa++ can be divided into three components corresponding to the time it takes to dispatch an action, the time required to synchronize using the pthreads library, and the time required to add an action to the scheduler via the ioa::schedule call.
From the results, we calculated the average overhead per action of the ioa++ framework when dispatching an action and found it to be 4316ns $\pm$ 3.5ns for the single threaded experiments and 6067ns $\pm$ 3.6ns for the multi-threaded experiments.
Similary, we calculated the scheduling overhead and found it to be 247ns $\pm$ 0.678ns for the single threaded experiments and 307ns $\pm$ 0.357ns for the multi-threaded experiments.
Both of these can be improved with better scheduler design.
The synchronization time depends on the interactions among the system being executed, the scheduler, and the locking scheme used by the framework.
We plan to address scheduling and synchronization in future work.

>>>>

The ioa++ framework permits independent actions to be executed concurrently.
The goal of this evaluation is to show that the degree of exploitable concurrency depends on the interactions of the I/O automata comprising a system and the overhead of the framework.
To test concurrent execution, we examine a system whose actions can be configured to span the range from having no independent actions to having only independent actions.
We measure the time required to execute a fixed number of actions using a scheduler with a configurable number of threads.
We then calculate the speed-up by comparing an execution using one thread to an equivalent execution using two threads.
We also vary the complexity of each action to gain insight into how overhead and synchronization affect concurrent execution.

An automaton $R$ contains an input, output, and internal action.
The output and internal action effects execute an algorithm whose complexity is proportional to the parameter $N$.
The automaton executes a fixed number of local actions.
The automaton schedules the internal action with probability $\sqrt{\rho}$ and schedules the output action otherwise.
The system $S$ to be executed consists of two $R$ automata composed so the output action of one is composed with the input action of the other.
If we divide the execution into rounds where each automaton executes a single action, the probability that both $R$ automata execute an internal action is $\sqrt{\rho} \times \sqrt{\rho} = \rho$.
Thus, we can use the parameter $\rho$ to vary the independence of the automata.

For this experiment, we developed a simple multi-threaded scheduler that assigns each local action to a thread based on the automaton identifier of the local action using the techniques described in Section~\ref{scheduling}.
When a single thread is used, all actions are executed by the same thread.
When two threads are used, one thread executes the actions of one $R$ automaton while the other thread executes the actions of the other $R$ automaton.

A trial consists of a choice for $\rho$, $N$, and the number of actions to be executed.
For our experiments, the parameter $\rho$ was varied from $0.0$ to $1.0$ in increments of $0.1$ and the parameter $N$ was varied from $1$ to $1000000$ by factors of $10$.
The number of actions executed by each automaton was fixed at 1000.
Each trial was repeated 1000 times.
All calculations were performed assuming a normal distribution at a confidence level of 95\%.

The driver for a single trial can be found in examples/random.cpp\footnote{Access directly at \url{http://github.com/jrwilson/ioa/blob/master/examples/random.cpp}.} of the ioa++ package.
The trials were performed on a Mac G4 running OS X 10.5.8 with two 2.66 GHz Dual-Core Intel Xeon processors with 2GB of memory.
The code was compiled and linked with i686-apple-darwin9-g++-4.0.1 build 5493 with -O2 optimization and the -DPROFILE flag to enable profiling in ioa++.

@@

Figure~\ref{speed_up} shows the speed-up when system $S$ was executed with one and two threads.
Confidence intervals are shown but are negligible with the largest being a speed-up factor of $\pm$0.0165.
When $\rho = 0$, every action is a bound output action and therefore depends on both $R$ automata.
Consequently, every action must be serialized yielding a maximum speed-up of 1.
When $\rho = 1$, every action is an internal action and can be executed concurrently with a corresponding maximum speed-up of 2.

@@

Figures~\ref{overhead1} and \ref{overhead2} show the average per thread of the fraction of time devoted to framework code and synchronization calls for the single and multi-threaded executions; confidence intervals are again negligible with largest being $\pm$0.002155 for Figure~\ref{overhead1} and $\pm$0.001781 for Figure~\ref{overhead2}.
Since the number of actions to be executed is constant, the single threaded execution shows a relatively consistent fraction for each value of $N$.
The fraction decreases as $\rho$ goes to 1 and the system transitions from acquiring two locks to one lock.
The multi-threaded execution shows that the fraction of time devoted to system overhead decreases as $\rho$ goes to 1 and is more pronounced when $N$ is large.
For example, when $N=1000000$ one thread spends half of its time waiting on the other thread when $\rho = 0$ and spends virtually no time on synchronization when $\rho = 1$.

This experiment indicates that concurrent execution is possible with ioa++ but a speed-up will only be achieved if (1) enough independent actions are enabled and (2) the duration of the independent actions overcome the overhead of ioa++.
The overhead of ioa++ can be divided into three components corresponding to the time it takes to dispatch an action, the time required to synchronize using the pthreads library, and the time required to add an action to the scheduler via the ioa::schedule call.
From the results, we calculated the average overhead per action of the ioa++ framework when dispatching an action and found it to be 4316ns $\pm$ 3.5ns for the single threaded experiments and 6067ns $\pm$ 3.6ns for the multi-threaded experiments.
Similary, we calculated the scheduling overhead and found it to be 247ns $\pm$ 0.678ns for the single threaded experiments and 307ns $\pm$ 0.357ns for the multi-threaded experiments.
Both of these potentially can be improved with better scheduler design.
The synchronization time depends on the interactions among the system being executed, the scheduler, and the locking scheme used by the framework.
We plan to address scheduling and synchronization in future work.
